---
title: "Fetal Endothelial Cells - Brain TF"
author: "Marina Nikolova"
date: "21/04/2022"
output: html_document
---

#### All Cao et al. 2021 datasets are in /links/groups/treutlein/USERS/zhisong_he/Work/public_datasets/Cao_Science_2020_fetal_human_atlas_scRNAseq/processed_datasets
#### There is the seurat_endothelial.RDS folder which contains the seurat object of all the endothelial cells
#### The seurat_endothelial_organs.RDS is a list of seurat objects with different separated by different organs
#### The multi-organ umap is in the all-cell object (with typical analysis without any integration), while the clusters in the heatmap are for each organ separately 
#### The script to process different organs separately is in reanalysis_endothelial/organ_separated_analysis.r

```{r setup, include=FALSE}

library(Matrix)
library(Seurat)
library(dplyr)
library(ggplot2)
library(readr)
library(viridis)
library(Rmagic)
library(simspec)
library(presto)
library(purrr)

# feature plots codes
source("/home/marinani/Scripts/Data_analysis/feature_plots.r")

# CSS script, Zhisong He
source("/home/marinani/Scripts/Data_analysis/CSS_integration_Zhisong.r")

# dimensioan reduction script, Zhisong He
source("/home/marinani/Scripts/Data_analysis/dimension_reduction_Zhisong.r")

```

### Load the list which contains the Seurat objects of all endothelial cells from Cao et al, separated by organ
```{r load the data, include = FALSE}

seurat_endo_orgs <- readRDS("/links/groups/treutlein/USERS/zhisong_he/Work/public_datasets/Cao_Science_2020_fetal_human_atlas_scRNAseq/processed_datasets/seurat_endothelial_organs.RDS")

```

### Explore the list
```{r general exploration}

seurat_endo <- readRDS("/links/groups/treutlein/USERS/zhisong_he/Work/public_datasets/Cao_Science_2020_fetal_human_atlas_scRNAseq/processed_datasets/seurat_endothelial.RDS")
cols_ages <- setNames(prettyrainbow_colscheme(length(unique(seurat_endo$Development_day))), sort(unique(seurat_endo$Development_day)))
#seurat_endo_orgs <- readRDS(file="~/Work/public_datasets/Cao_Science_2020_fetal_human_atlas_scRNAseq/processed_datasets/seurat_endothelial_organs.RDS")
layout(matrix(1:15,nrow=3,byrow=T)); par(mar=c(1,1,3,1))
for(org in names(seurat_endo_orgs)){
  plotFeature(Embeddings(seurat_endo_orgs[[org]], "umap"), factor(seurat_endo_orgs[[org]]$Development_day), colorPal=cols_ages, cex=0.6, main=org, cex.main=1.5)
}
plot(1,1,type="n",bty="n", axes=F); legend("topleft", legend=names(cols_ages)[1:10], pch=16, col=cols_ages[1:10], bty="n", cex=1.2); legend("top", legend=names(cols_ages)[-(1:10)], pch=16, col=cols_ages[-(1:10)], bty="n", cex=1.2)

plotMultiFeatures(Embeddings(seurat_endo_orgs$Adrenal,"umap"), seurat_endo_orgs$Adrenal@assays$RNA@data[c("CYP21A2","CYP11A1","NCK2","NPIPP1","ANKRD29","SNX25","HMBS","RP11-540K16.1"),],
                  colorPal=beach_colscheme, cex=0.6, random_order=F, sort_by_value=T, cex.main=1.5, ncol=4)

FeaturePlot(seurat_endo_orgs$Lung, c("PLVAP"), reduction="umap", cols=beach_colscheme(30), order=T) & NoAxes() & NoLegend()
DimPlot(seurat_endo_orgs$Cerebellum, reduction="umap")

```

### Now we merge the objects
```{r merge the objects}

# create a function to merge the objects (seurat provides only a function that requires at least one additional separate object)
Merge_Seurat_List <- function(
  list_seurat,
  add.cell.ids = NULL,
  merge.data = TRUE,
  project = "SeuratProject"
) {
  merged_object <- reduce(list_seurat, function(x, y) {
    merge(x = x, y = y, add.cell.ids = add.cell.ids, merge.data = merge.data, project = project)
  })
}

merged_seurat_endo_orgs <- Merge_Seurat_List(
  seurat_endo_orgs,
  add.cell.ids = NULL,
  merge.data = TRUE,
  project = "endothelial_organs"
)

saveRDS(merged_seurat_endo_orgs, file="/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/Brain_Vasculature/fetal_brain_endo_cells/merged_seurat_endo_orgs.rds")
# merged_seurat_endo_orgs <- readRDS("/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/Brain_Vasculature/fetal_brain_endo_cells/merged_seurat_endo_orgs.rds")

```
# Find all markers per organ
```{r find all markers, echo=FALSE}

Idents(merged_seurat_endo_orgs) <- "Organ"
all_markers <- FindAllMarkers(merged_seurat_endo_orgs, only.pos = FALSE, min.pct = 0.1, logfc.threshold = log(0.25))
all_markers %>%
    group_by(cluster) %>%
    slice_max(n = 30, order_by = avg_log2FC)

write.csv(all_markers, "/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/Brain_Vasculature/fetal_brain_endo_cells/all_markers_merged_seurat_endo_orgs.csv")

```

### Load the list with human-specific transcription factors
##### The list has been downloaded from Animal TFDB v3 (http://bioinfo.life.hust.edu.cn/AnimalTFDB/)
##### It contains 1665 TFs and 1025 TF Cofactors for Homo sapiens
```{r load the human TF list}

human_TF <- read.table("/home/marinani/Scripts/Databases/AnimalTFDB/AnimalTFDB_v3/Homo_sapiens_TF.tsv", sep = '\t', header = TRUE, fill = TRUE)

```

### Generate a data frame from all amrkers which contains the TFs only
```{r extract TFs}

all_markers_filtered <- as.data.frame(all_markers) %>% 
      filter(gene %in% human_TF$Symbol)  %>% 
      arrange(cluster,desc(avg_log2FC))
all_markers_filtered 

# add a column for no, up or down regulation
all_markers_filtered$diffexpressed <- "No"
# define up regulated as lo2FC > 0.5, pvalue < 0.05
all_markers_filtered$diffexpressed[all_markers_filtered$avg_log2FC > 0.5 & all_markers_filtered$p_val_adj < 0.05] <- "Up"
# define down regulated as lo2FC < -0.5, pvalue < 0.05
all_markers_filtered$diffexpressed[all_markers_filtered$avg_log2FC < -0.5 & all_markers_filtered$p_val_adj < 0.05] <- "Down"

# add label to DEGs
all_markers_filtered$delabel <- NA
all_markers_filtered$delabel[all_markers_filtered$diffexpressed != "No"] <- all_markers_filtered$gene[all_markers_filtered$diffexpressed != "No"]

write.csv(all_markers_filtered, "/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/Brain_Vasculature/fetal_brain_endo_cells/all_markers_filtered.csv")

```

### Plot the TFs with hight log fold change in cerebrum and cerebellum fetal endothelial cells against the difference of cells in other organs expressing it (in percentage)
```{r plot}

library(ggrepel)

cerebellum <- all_markers_filtered %>%
  filter(cluster == "Cerebellum") %>%
  ggplot(aes(x=avg_log2FC,y=pct.1-pct.2, col=diffexpressed, label=delabel)) +
  geom_point() +
  geom_text_repel() +
  theme_minimal() +
  ggtitle("Cerebellum") +
  theme(plot.title = element_text(size=15, hjust=0.5))
cerebellum

cerebrum <- all_markers_filtered %>%
  filter(cluster == "Cerebrum") %>%
  ggplot(aes(x=avg_log2FC,y=pct.1-pct.2, col=diffexpressed, label=delabel)) +
  geom_point() +
  geom_text_repel() +
  theme_minimal() +
  ggtitle("Cerebrum") +
  theme(plot.title = element_text(size=15, hjust=0.5))
cerebrum

```

